---
title: "BIOL 432 — Assignment 8: Dragon Phylogeny"
author: "Justin Lasrado"
date: "2025-10-28"
output: html_document
---

## Overview

For this assignment I used the Dragon Phylogeny character matrix from the Colautti Lab and added three dragons from *Game of Thrones*: Drogon, Rhaegal, and Viserion. I then calculated distances between the dragons and built a Neighbor-Joining tree.

The original Dragon Phylogeny project uses binary-coded traits. Missing traits are represented by `?` in the Nexus file.

## Packages

```{r setup, message=FALSE, warning=FALSE}
library(ape)
library(ggplot2)
library(ggtree)

knitr::opts_chunk$set(echo = TRUE)
```

## Read in the original data

```{r read-data}
DragonNexus <- read.nexus.data("Data/DragonMatrix.nex")

# Number of traits in the dataset
num_traits <- length(DragonNexus[[1]])

DragonNexusDF <- data.frame(
  matrix(unlist(DragonNexus), ncol = num_traits, byrow = TRUE),
  stringsAsFactors = FALSE
)

row.names(DragonNexusDF) <- names(DragonNexus)

head(DragonNexusDF[, 1:5])
dim(DragonNexusDF)
```

The original matrix should contain 77 dragons and 78 binary characters.

```{r check-data}
stopifnot(nrow(DragonNexusDF) == 77)
stopifnot(ncol(DragonNexusDF) == 78)
```

## Add the three dragons

The three dragons were coded using the character values from the assignment. The remaining characters are coded as `0` so that each new dragon has the same 78-character structure as the original matrix.

```{r add-dragons}
# Add the three dragons
new_codes <- matrix(
  c(
    # Drogon
    "004","005","002","002","003","001","004","003","002","002","003","001","002","002","004","002","001","005","001",
    # Rhaegal
    "004","004","002","002","002","002","003","002","002","002","002","001","002","002","004","002","001","004","001",
    # Viserion
    "004","004","003","002","002","001","004","004","003","003","004","001","003","002","004","002","002","004","001"
  ),
  nrow = 3,
  byrow = TRUE
)

# Fill the remaining characters with 0
new_codes_padded <- cbind(
  new_codes,
  matrix("000", nrow = 3, ncol = num_traits - ncol(new_codes))
)

new_dragons <- data.frame(new_codes_padded, stringsAsFactors = FALSE)
row.names(new_dragons) <- c("Drogon", "Rhaegal", "Viserion")
colnames(new_dragons) <- colnames(DragonNexusDF)

nexus_updated <- rbind(DragonNexusDF, new_dragons)

# Make sure the new matrix has the expected dimensions
stopifnot(nrow(nexus_updated) == 80)
stopifnot(ncol(nexus_updated) == 78)

head(nexus_updated)
tail(nexus_updated, 3)
```

Save the updated matrix so it can be checked separately from the original data.

```{r save-data}
dir.create("input", showWarnings = FALSE)
write.csv(
  nexus_updated,
  "input/nexus_with_dragons.csv",
  row.names = TRUE,
  quote = TRUE
)
```

## Distance matrix

The character matrix contains `0`, `1`, and `?` values. For the distance calculation, I convert the observed binary characters to numeric values and leave missing values as `NA`.

```{r distance}
# Convert the character matrix to numeric values
DragonNumeric <- matrix(
  as.numeric(ifelse(nexus_updated == "?", NA, nexus_updated)),
  nrow = nrow(nexus_updated),
  ncol = ncol(nexus_updated),
  dimnames = dimnames(nexus_updated)
)

DragonDist <- dist(DragonNumeric, method = "binary")
DragonDistMatrix <- as.matrix(DragonDist)

# Check the distance matrix
round(DragonDistMatrix[1:5, 1:5], 3)
```

The diagonal should be zero because each dragon has a distance of zero from itself.

```{r distance-check}
stopifnot(all(diag(DragonDistMatrix) == 0))
```

## Plot the distance matrix

```{r distance-plot}
PDat <- as.data.frame(as.table(DragonDistMatrix))
names(PDat) <- c("Query", "Subject", "Distance")

ggplot(PDat, aes(x = Query, y = Subject, fill = Distance)) +
  geom_tile() +
  scale_fill_viridis_c(option = "magma") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, size = 5),
    axis.text.y = element_text(size = 5)
  ) +
  labs(x = NULL, y = NULL, fill = "Distance")
```

## Build the Neighbor-Joining tree

```{r tree}
DragonTree <- nj(DragonDist)

# Check that all 80 dragons are represented in the tree
stopifnot(length(DragonTree$tip.label) == 80)

DragonTree
```

## Colour the tree by the names in the original dataset

The names in the matrix include information about where the original dragon image came from. I grouped those names into broad categories for the tree visualization.

```{r groups}
Traits <- data.frame(
  Name = DragonTree$tip.label,
  group = NA_character_,
  stringsAsFactors = FALSE
)

Traits$group[grep("German", Traits$Name, ignore.case = TRUE)] <- "Western"
Traits$group[grep("French", Traits$Name, ignore.case = TRUE)] <- "Western"
Traits$group[grep("Dutch", Traits$Name, ignore.case = TRUE)] <- "Western"
Traits$group[grep("English", Traits$Name, ignore.case = TRUE)] <- "Western"
Traits$group[grep("American", Traits$Name, ignore.case = TRUE)] <- "Western"
Traits$group[grep("British", Traits$Name, ignore.case = TRUE)] <- "Western"
Traits$group[grep("Spanish", Traits$Name, ignore.case = TRUE)] <- "Western"
Traits$group[grep("Italian", Traits$Name, ignore.case = TRUE)] <- "Western"
Traits$group[grep("Japanese", Traits$Name, ignore.case = TRUE)] <- "Orientalia"
Traits$group[grep("Indian", Traits$Name, ignore.case = TRUE)] <- "Orientalia"
Traits$group[grep("Iranian", Traits$Name, ignore.case = TRUE)] <- "Orientalia"
Traits$group[grep("Greece", Traits$Name, ignore.case = TRUE)] <- "Ancient"
Traits$group[grep("Turkish", Traits$Name, ignore.case = TRUE)] <- "Ancient"
Traits$group[grep("Snake", Traits$Name, ignore.case = TRUE)] <- "Serpentidae"
Traits$group[grep("Fish", Traits$Name, ignore.case = TRUE)] <- "Aquatic"
Traits$group[grep("Mammal", Traits$Name, ignore.case = TRUE)] <- "Beast"
Traits$group[is.na(Traits$group)] <- "Other"

Traits$group <- factor(Traits$group)
table(Traits$group)
```

## Final tree

```{r final-tree, fig.width=10, fig.height=10}
ggtree(DragonTree, layout = "circular") %<+% Traits +
  geom_tiplab(aes(color = group), size = 2.2, show.legend = TRUE) +
  scale_color_manual(values = c(
    "Western" = "red",
    "Orientalia" = "blue",
    "Ancient" = "lightblue",
    "Serpentidae" = "green",
    "Aquatic" = "black",
    "Beast" = "purple",
    "Other" = "gray"
  )) +
  ggtitle("Dragon Phylogeny") +
  theme_tree2() +
  theme(
    legend.position = "right",
    legend.text = element_text(size = 8),
    legend.title = element_blank(),
    plot.title = element_text(hjust = 0.5)
  )
```

## Notes

This tree is based on the character matrix and the Neighbor-Joining method. The three new dragons are added to the same matrix as the original 77 observations, so they are included in the same distance calculation and tree.

The original Dragon Phylogeny dataset and tutorial were provided through the Colautti Lab repository. The original dataset is being used here as course material; the analysis and additions above are my Assignment 8 work.
